Project 2

### 2018-10-01 Project2/report/Project2.tex
Finishing up method and results

\subsection{Buckling beam}
Considering a beam that is fastened in both ends which moves up and down when a force F is exerted on it can be described by a differential equation
%
\begin{gather}\label{eq:du2/dx2}
    R \frac{d^2 u(x)}{dx^2} = -Fu(x),
\end{gather}
%
where $u(x)$ is the displacement of the beam in the y-direction and $R$ is a constant that is defined by the properties of the beam, such as the rigidity. Since the beam, with length L, is fastened in both ends, the displacement at the end points $u(0) = u(L) = 0$, so we have Dirichlet boundary condition as in the previous paper \cite{1}, and the sum of outer forces exerted on the beam is zero. Introducing the dimensionless variable $\rho = x/L \Rightarrow x = \rho L$, where $\rho \in [0,1]$. Introducing this into Eq. \eqref{eq:du2/dx2} gives
%
\begin{gather*}
    \frac{R}{L}\frac{d^2 u(\rho)}{d\rho^2} = - FLu(\rho)
    \Rightarrow \frac{d^2 u(\rho)}{d\rho^2} = -\frac{FL^2}{R}u(\rho).
\end{gather*}
%
Substituting $\lambda = FL^2/R$ transforms the problem to an eigenvalue problem, with discretized $\lambda$'s as eigenvalues with corresponding eigenvectors. The methods presented in the previous paper regarding the solution of a differential equation will be applied to this problem. The left hand size of the equation to solve,
%
\begin{gather}\label{eq:du/dx=lambda}
    \frac{d^2 u(\rho)}{dx^2} = -\lambda u(\rho)
\end{gather}
%
can be approximated as
%
\begin{gather*}
    \frac{d^2 u(\rho)}{dx^2} = \frac{u(\rho + h) - 2u(\rho) + u(\rho - h)}{h^2} +  \mathcal{O}(h^2),
\end{gather*}
%
where the step size $h = \left(\rho_{max} - \rho_{min}\right)/n$, which is the total length of the beam divided by the resolution of the calculations n. The maximum and minimum value of the dimensionless length variable $\rho$ is at the end points of the beam, such that $\rho_{max} = 1$ and $\rho_{min} = 0$. Following the method from the previous paper the discretized equation to solve is
%
\begin{gather*}
    -\frac{u_{i+1} - 2u_i + u_{i-1}}{h^2} = \lambda u_i.
\end{gather*}
Which by the method of the previous paper can be transferred to a matrix eigenvalue problem;
%
\begin{equation*}
    \begin{bmatrix} d& a & 0   & 0    & \dots  &0     & 0 \\
                                a & d & a & 0    & \dots  &0     &0 \\
                                0   & a & d & a  &0       &\dots & 0\\
                                \dots  & \dots & \dots & \dots  &\dots      &\dots & \dots\\
                                0   & \dots & \dots & \dots  &a  &d & a\\
                                0   & \dots & \dots & \dots  &\dots       &a & d\end{bmatrix} 
                                 \begin{bmatrix} u_1 \\ u_2 \\ u_3 \\ \dots \\ u_{N-2} \\ u_{N-1}\end{bmatrix} = \lambda \begin{bmatrix} u_1 \\ u_2 \\ u_3 \\ \dots \\ u_{N-2} \\ u_{N-1}\end{bmatrix} . 
\label{eq:matrixse} 
\end{equation*}
%
Where a and b are defined as $a = -1/h^2$ and $b = 2/h^2$. For this specific eigenvalue problem the analytical eigenvalues are
%
\begin{gather}\label{eq:eigenvals_analytical}
    \lambda_j = d + 2a\cos\left(\frac{j\pi}{n+1}\right)
\end{gather}
for $j = 1,2,...,n-1$.

@@

@@

Comparing the results from Jacobi's method and the analytic results from the provided in [SOURCE:Poject 2], we achieve sufficient numerical precision even for variyng sizes between $n = 5$ and $n = 400$.
%
\begin{deluxetable}{lcc}
%\tablewidth{0pt}
\tablecaption{\label{tab:results}}
\tablecomments{Table comparing CPU-time between Armadillo's built in function (eigsym) and Jacobi's method, as visible from this data Jacobi's method increases in time spent considerably faster than that of Armadillo's function.}
\tablecolumns{4}
\tablehead{Size n  & Armadillo & Jacobi's method}
\startdata
10 & 0.000262s & 0.000155s \\
50 & 0.000915s & 0.007585s \\
100 & 0.005636s & 0.106488s \\
200 & 0.018883s & 1.53787s \\
300  & 0.04241s & 8.85018s \\
400  & 0.085901s & 30.4565s
\enddata
\end{deluxetable}
%
%
\begin{figure}[t]
\mbox{\epsfig{figure=iterationspersize15.pdf,width=\linewidth,clip=}}
\caption{Logarithmic plot of the number of iteration needed as function of the matrix dimentionality n, to illustrate the quadratic dependency we have plotted $y = 2log_{10}(n)$, which in the plot is seen as the dotted $y=2x$ line.}
\label{fig:fig1}
\end{figure}
%
\subsection{Quantum dots, single electron}
Analyzing how the number of integration points $n$ and the endpoint $\rho_{n}$, and how many integration points is needed to achieve four digit accuracy in the eigenvalues as a result from using Jacobi's method on a Toepliz matrix with quantum harmonic oscillator potential term. For $\rho_{n} = 5$, the size n required to achieve four leading digits is $n = 380$, which returns the value $E_{00} = 2.99995$ .

@@

>>>>

\subsection{Buckling beam}
Considering a beam that is fastened in both ends which moves up and down when a force F is exerted on it can be described by a differential equation
%
\begin{gather}\label{eq:du2/dx2}
    R \frac{d^2 u(x)}{dx^2} = -Fu(x),
\end{gather}
%
where $u(x)$ is the displacement of the beam in the y-direction and $R$ is a constant that is defined by the properties of the beam, such as the rigidity. Since the beam, with length L, is fastened in both ends, the displacement at the end points $u(0) = u(L) = 0$, so we have Dirichlet boundary condition as in the previous paper \cite{1}, and the sum of outer forces exerted on the beam is zero. Introducing the dimensionless variable $\rho = x/L \Rightarrow x = \rho L$, where $\rho \in [0,1]$. Introducing this into Eq. \eqref{eq:du2/dx2} gives
%
\begin{gather*}
    \frac{R}{L}\frac{d^2 u(\rho)}{d\rho^2} = - FLu(\rho)
    \Rightarrow \frac{d^2 u(\rho)}{d\rho^2} = -\frac{FL^2}{R}u(\rho).
\end{gather*}
%
Substituting $\lambda = FL^2/R$ transforms the problem to an eigenvalue problem, with discretized $\lambda$'s as eigenvalues with corresponding eigenvectors. The methods presented in the previous paper regarding the solution of a differential equation will be applied to this problem. The left hand size of the equation to solve,
%
\begin{gather}\label{eq:du/dx=lambda}
    \frac{d^2 u(\rho)}{dx^2} = -\lambda u(\rho)
\end{gather}
%
can be approximated as
%
\begin{gather*}
    \frac{d^2 u(\rho)}{dx^2} = \frac{u(\rho + h) - 2u(\rho) + u(\rho - h)}{h^2} +  \mathcal{O}(h^2),
\end{gather*}
%
where the step size $h = \left(\rho_{max} - \rho_{min}\right)/n$, which is the total length of the beam divided by the resolution of the calculations n. The maximum and minimum value of the dimensionless length variable $\rho$ is at the end points of the beam, such that $\rho_{max} = 1$ and $\rho_{min} = 0$. Following the method from the previous paper the discretized equation to solve is
%
\begin{gather*}
    -\frac{u_{i+1} - 2u_i + u_{i-1}}{h^2} = \lambda u_i.
\end{gather*}
Which by the method of the previous paper can be transferred to a matrix eigenvalue problem;
%
\begin{equation*}
    \begin{bmatrix} d& a & 0   & 0    & \dots  &0     & 0 \\
                                a & d & a & 0    & \dots  &0     &0 \\
                                0   & a & d & a  &0       &\dots & 0\\
                                \dots  & \dots & \dots & \dots  &\dots      &\dots & \dots\\
                                0   & \dots & \dots & \dots  &a  &d & a\\
                                0   & \dots & \dots & \dots  &\dots       &a & d\end{bmatrix} 
                                 \begin{bmatrix} u_1 \\ u_2 \\ u_3 \\ \dots \\ u_{N-2} \\ u_{N-1}\end{bmatrix} = \lambda \begin{bmatrix} u_1 \\ u_2 \\ u_3 \\ \dots \\ u_{N-2} \\ u_{N-1}\end{bmatrix} . 
\label{eq:matrixse} 
\end{equation*}
%
Where a and b are defined as $a = -1/h^2$ and $b = 2/h^2$. For this specific eigenvalue problem the analytical eigenvalues are, as stated in \cite{2}
%
\begin{gather}\label{eq:eigenvals_analytical}
    \lambda_j = d + 2a\cos\left(\frac{j\pi}{n+1}\right)
\end{gather}
for $j = 1,2,...,n-1$.

@@

\subsection{Quantum dots, two electrons}
Consider now two electrons in a harmonic oscillator potential. The electrons will interact through a repulsive Coulomb intereaction. For two non-interacting particles, the radial part of the Schr$\ddot{\mathrm{o}}$dinger equation is
%
\begin{equation*}
    \left[\frac{-\hbar^2}{2m}\left[\frac{d^2}{dr_1^2} - \frac{d^2}{dr_2^2}\right] + \frac{k\left(r_1^2 + r_2^2\right)}{2}\right]u(r_1,r_2) = E^{(2)}u(r_1,r_2),
\end{equation*}
%
where $r_1, r_2$ denotes the radii from the center of the harmonic oscillator potential to electron 1 and 2 respectively. The superscript of the energy corresponds to the energy of the two electrons, such that the energy of the single electron previously discussed would be denoted as $E^{(1)}$. Consider a relative coordinate $\bold{r} = \bold{r_1} - \bold{r_2}$ and a center of mass coordinate $\bold{R} = \left(\bold{r_1} + \bold{r_2}\right)/2$, which transforms the Schr$\ddot{\mathrm{o}}$dinger equation to
%
\begin{equation*}
    \left[\frac{-\hbar^2}{m}\left[\frac{d^2}{dr^2} - \frac{1}{4}\frac{d^2}{dR^2}\right] + \frac{kr^2}{4} + kR^2\right]u(r,R) = E^{(2)}u(r,R)
\end{equation*}
%
where the energy is now $E^{(2)} = E_r + E_R$.  Adding the Coulomb repulsion corresponds to adding a potential term
%
\begin{equation*}
    V(r_1,r_2) = \frac{e^2}{4\pi \epsilon_0 |\bold{r_1} - \bold{r_2}|} = \frac{\beta e^2}{r}
\end{equation*}
%
where $\epsilon_0$ is the permittivity of vacuum, and $\beta e^2 = 1.44$ eVnm.

@@

Comparing the results from Jacobi's method and the analytic results from the provided in [SOURCE:Poject 2], we achieve sufficient numerical precision even for variyng sizes between $n = 5$ and $n = 400$.
%
\begin{deluxetable}{lcc}
%\tablewidth{0pt}
\tablecaption{\label{tab:results}}
\tablecomments{Table comparing CPU-time between Armadillo's built in function \textbf{eigsym} and Jacobi's method, as visible from this data Jacobi's method increases in time spent considerably faster than that of Armadillo's function.}
\tablecolumns{4}
\tablehead{Size n  & Armadillo & Jacobi's method}
\startdata
10 & 0.000262s & 0.000155s \\
50 & 0.000915s & 0.007585s \\
100 & 0.005636s & 0.106488s \\
200 & 0.018883s & 1.53787s \\
300  & 0.04241s & 8.85018s \\
400  & 0.085901s & 30.4565s
\enddata
\end{deluxetable}
%
%
\begin{figure}[t]
\mbox{\epsfig{figure=iterationspersize15.pdf,width=\linewidth,clip=}}
\caption{Logarithmic plot of the number of iteration needed as function of the matrix dimentionality n, to illustrate the quadratic dependency we have plotted $y = 2log_{10}(n)$, which in the plot is seen as the dotted $y=2x$ line.}
\label{fig:fig1}
\end{figure}
%
\subsection{Quantum dots, single electron}
Analyzing how the number of integration points $n$ and the endpoint $\rho_{n}$, and how many integration points is needed to achieve four digit accuracy in the eigenvalues as a result from using Jacobi's method on a Toepliz matrix with quantum harmonic oscillator potential term. For $\rho_{n} = 5$, the size n required to achieve four leading digits is $n = 380$, which returns the value $E_{00} = 2.99995$, where the analytic eigenvalue is $\lambda = 3$.

@@

\bibitem{2}
Morten Hjorth Jensen
\textit{Project 2}
\url{http://compphysics.github.io/ComputationalPhysics/doc/Projects/2018/
Project2/pdf/Project2.pdf} (01.10.18)
